Wait-free Queues with Polygarithmic Step Complexity

### 2022-06-16 article/psuedocode.tex
add lemmas, remove helping, apply comments

\begin{algorithm}
\caption{Fields description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{100}
\begin{multicols}{2}

@@

\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
\begin{itemize}

  \item \tt{\sl{int} \tt{last\sub{done}}}

  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the. A block is finished if all of its operations are finished. \tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response. \it{put the definitions before the pseudocode}}

\end{itemize}
\pagebreak

@@

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} length}
  \textsf{: length of the queue after performing all operations in the prefix for this block}
%  \item \tt{\sl{int} sum\sub{non-null deq}}
%  \textsf{: count of non-null dequeus up to this block}
  \item \tt{\sl{counter} num\sub{finished}}
  \textsf{: number of finished operations in the block}
%    \item \tt{\sl{int} order}
%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
\end{itemize}

@@

@@

\Statex
\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \Return\tt{leaf.}\Call{HelpDequeue()}{}
\EndFunction{Dequeue}

\Statex
\pagebreak

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
\State \Return \tt{<root.\Call{BSearch}{sum\sub{enq}, r\sub{enq}, root.FindMostRecentDone(), root.size}, r\sub{enq}>}
\EndIf
\EndFunction{FindResponse}

@@

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.size} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{this\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{500}

\Function{Block}{FindMostRecentDone}{}
\For{\tt{leaf l} \bf{in leaves}}
\State\tt{max= Max(l.maxOld, max)}
\EndFor
\State\Return\tt{max} \cmt{This snapshot suffies.}
\EndFunction{FindMostRecentDone}

\end{algorithmic}
\end{algorithm}
%##########################################

@@

\Function{Object}{HelpDequeue()}{}
\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
 \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist. If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
\State \tt{last\sub{done}= b\sub{deq}}
\EndIf
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
\State \tt{last\sub{done}= b\sub{deq}}
\ElsIf{\tt{root.blocks[b\sub{enq}].num\sub{finished}==root.blocks[b\sub{enq}].num}}
\State \tt{last\sub{done}= b\sub{enq}}
\EndIf
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

\Function{void}{Help}{}\cmt{Helps pending operations}

\State{\tt{last= l.size-1}}
\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
\EndIf

\EndFunction{Help}

@@

\Statex
\Statex $\diamondsuit$ \tt{\sl{root implementation}}
%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
\State \tt{l.Help()}
\EndFor
\EndIf
\State \tt{blk.num\sub{finished}= 0}
%\State \tt{*oldRoot= \&root.blocks.root}
%\State \tt{*newRoot= root.blocks.Append(blk).root}
%\State \Return \tt{CAS(root, oldRoot, newRoot)}
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

\begin{algorithm}
\caption{Yet to decide how to handle.}
\begin{algorithmic}[1]
\setcounter{ALG@line}{800}

%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
%\State \tt{t1,t2= RBT.split(order, s)}
%\State \tt{RBTRoot.CAS(t2.root)}
%\EndFunction{CollectGarbage}

\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}

\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}

\State \Return \tt{this.leaf.response(block.order)}
\EndIf

\EndFunction{FallBack}

@@

\clearpage
\section{Proof of Linearizability}

@@

\begin{definition}[Block]
A block is an object storing some statistics described in Algorithm Queue. It implicitly shows a set of operations. The set of operations of block $b$ are the operations in the leaf subblocks of $b$. We show the set of operations of block $b$, set of blocks $B$ by $ops(b)$, $ops(B)$. We also say $b$ contains $op$ if $op\in ops(b)$.
\end{definition}

\begin{definition}[Order]
If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}. Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s. 
\end{definition}

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it. Block b is direct superblock of block c if c is direct subblock of b.
\end{definition}

@@

\begin{definition}
  Block \nf{b} in \nf{n.blocks} is \emph{Established} at time $t$ if \nf{n.head} is greater than index of \nf{b} at time $t$. Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
\end{definition}

\begin{observation}\label{head}
Once block b is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
\end{observation}

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left, right}} $\geq$ \nf{n.blocks[i-1].end\sub{left, right}}.
\end{lemma}
\begin{proof}
  Induced trivially from the pseudocode since \nf{n.head} is only incremented. \nf{n.blocks[i]. Also end\sub{left, right}} are greater than or equal the previous values in the \nf{CreateBlock(i)} code (Lines~\ref{lastLine, prevLine}).
\end{proof}

\begin{lemma}
Every block has most one direct superblock.
\end{lemma}
\begin{proof}
To show this we are going to refer to the way n.blocks[] is partitioned while propagating up to n.parent. \nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}). From we know that \nf{end\sub{left},end\sub{right}} are non decreasing, so the range of the subblocks of n.blocks are disjoint.
\end{proof}

\begin{corollary}[No Duplicates]\label{append}
If \nf{op} is appended to \nf{n.blocks[i]} then after that there is no \nf{j>i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
\end{corollary}
%\begin{proof}
%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
%\end{proof}

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
\end{invariant}
\begin{proof}
Af first the invariant is true since initially 1 is assigned to \nf{n.head} and \nf{n.blocks[x]} is null for every \nf{x}. The satisfiability of the invariant may be effected by writing into \nf{n.blocks} or incrementing \nf{n.head}.

Some value is written into \nf{n.blocks[head]} only in Line 313. It is obvious that writing into \nf{n.blocks[head]} does not change the state of the claim of the lemma. The value of \nf{n.head} is modified only in lines \ref{incrementHead1}, ][p\ref{incrementHead2}, we show in both cases of the successful \ref{cas} or not the claim holds after the increment lines of \nf{n.head}. If $h$ increments to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to claim the invariant still holds. In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore. Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readSize}. The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
%At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}

@@

\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
  If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
\end{lemma}
\begin{proof}
Blocks are only appended(not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node grows.
\end{proof}

@@

\nf{CreateBlock()} reads blocks in the children that do not exist in the parent and aggregates them into one block. If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence. So suppose two \nf{Refresh}es fail. Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}. So it means the second failed \nf{Refresh} is concurrent with a successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}. After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.

>>>>

\begin{algorithm}
\caption{Tree Fields Description}
\begin{algorithmic}[1]
\setcounter{ALG@line}{100}
\begin{multicols}{2}

@@

%\Statex $\blacktriangleright$ \tt{\sl{Leaf} extends \sl{Node}}
%\begin{itemize}
%  
%  \item \tt{\sl{int} \tt{last\sub{done}}}
%
%  \textsf{Stores the index of the block in the root such that the process that owns this leaf has most recently finished the. A block is finished if all of its operations are finished. \tt{enqueue(e)} is finished if \tt{e} is returned by some \tt{dequeue()} and \tt{dequeue()} is finished when it computes its response. \it{put the definitions before the pseudocode}}
%  
%\end{itemize}

@@

\Statex $\blacktriangleright$ \tt{\sl{RootBlock} extends \sl{InternalBlock}}
\begin{itemize}
  \item \tt{\sl{int} length}
  \textsf{: length of the queue after performing all operations in the prefix for this block}
%  \item \tt{\sl{int} sum\sub{non-null deq}}
%  \textsf{: count of non-null dequeus up to this block}
%  \item \tt{\sl{counter} num\sub{finished}}
%  \textsf{: number of finished operations in the block}
%    \item \tt{\sl{int} order}
%  \textsf{: the index of the block in the \tt{BlockList} containing the block.}
\end{itemize}

@@

%\Statex
%\Statex $\diamondsuit$ \tt{\sl{root implementation}}
%%  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
%%  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
%%\Statex \tt{root: }\sf{pointer to the root of the PBRT}
%\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
%\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
%\For{\tt{leaf l} \bf{in} \tt{tree leaves}}
%\State \tt{l.Help()}
%\EndFor
%\EndIf
%\State \tt{blk.num\sub{finished}= 0}
%%\State \tt{*oldRoot= \&root.blocks.root}
%%\State \tt{*newRoot= root.blocks.Append(blk).root}
%%\State \Return \tt{CAS(root, oldRoot, newRoot)}
%\State \Return{\tt{CAS(blocks[n], null, blk)}}
%\EndFunction{TryAppend}

@@

\Function{<int, int>}{FindResponse}{\sl{int} b, \sl{int} i}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
\State \Return \tt{<root.\Call{DSearch}{r\sub{enq}, b}, r\sub{enq}>}
\EndIf
\EndFunction{FindResponse}

\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
\State \tt{<b\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{b\sub{deq}, r\sub{deq}} 
 \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist. If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

\end{multicols}
\end{algorithmic}
\end{algorithm}
\pagebreak

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{root.blocks[end].sum\sub{enq} $\geq$ \tt{r\sub{enq}}}}
\Function{int}{DSearch}{\sl{int} i, \sl{int} end}

\Statex \cmt{\textmd{Searches for~the \tt{i}th enqueue of the given prefix sum of \tt{b}th block in the \tt{root}. Returns the index of the leftmost block in \tt{root.blocks} whose \sl{sum\sub{enq}} is $\geq$ \tt{i}}.}
\State \tt{start= b-1}
\While{\tt{root.blocks[start].sum\sub{enq}}$\geq$\tt{r\sub{enq}}}
\State \tt{start= start-(b-start)}
\EndWhile
\Return \tt{root.BSearch(sum\sub{enq}, r\sub{enq}, start, b)}
\EndFunction{DSearch}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.size} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

\If{\tt{this} \bf{is} \tt{root}}
\State \tt{newBlock.length= max(root.blocks[i-1].length + b.num\sub{enq} - b.num\sub{deq}, 0)}\label{computeLength}
%\State \tt{b.sum\sub{non-null deq}= root.blocks[i-1].sum\sub{non-null deq} + max( b.num\sub{deq} - root.blocks[i-1].length - b.num\sub{enq}, 0)}
\EndIf

@@

%\begin{algorithm}
%\caption{Root}
%\begin{algorithmic}[1]
%\setcounter{ALG@line}{500}
%
%\Function{Block}{FindMostRecentDone}{}
%\For{\tt{leaf l} \bf{in leaves}}
%\State\tt{max= Max(l.maxOld, max)}
%\EndFor
%\State\Return\tt{max} \cmt{This snapshot suffies.}
%\EndFunction{FindMostRecentDone}
%
%\end{algorithmic}
%\end{algorithm}
%##########################################

@@

%\Statex
%
%\Function{void}{Help}{}\cmt{Helps pending operations}
%
%\State{\tt{last= l.size-1}}
%\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
%\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
%\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
%\EndIf
%
%\EndFunction{Help}

@@

%\begin{algorithm}
%\caption{Yet to decide how to handle.}
%\begin{algorithmic}[1]
%\setcounter{ALG@line}{800}
%
%%\Function{void}{CollectGarbage}{}\cmt{Collects the root blocks that are done.}
%%\State \tt{s=FindMostRecentDone(Root.Blocks.root)}  \cmt{Lemma: If block b is done after helping then all blocks before b are done as well.}
%%\State \tt{t1,t2= RBT.split(order, s)}
%%\State \tt{RBTRoot.CAS(t2.root)}
%%\EndFunction{CollectGarbage}
%
%\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
%
%\If{root.blocks.get(num\sub{enq}), i is null} \cmt{this enqueue was already finished}
%
%\State \Return \tt{this.leaf.response(block.order)}
%\EndIf
%
%\EndFunction{FallBack}
%
%\end{algorithmic}
%\end{algorithm}

@@

\clearpage

\section{Proof of Linearizability}

@@

\begin{definition}[Block]
A block is an object storing some statistics, as described in Algorithm Queue. It implicitly represents a set of operations. If \nf{n.blocks[i]==b}  we call \nf{i} the \emph{index} of block \nf{b}. Block \nf{b} is before block \nf{b$^\prime$} in node \nf{n} if and only if \nf{b}'s index is smaller than \nf{b$^\prime$}'s. 
\end{definition}

\begin{definition}[Subblock]
Block \texttt{b} is a \emph{direct subblock} of \texttt{n.blocks[i]} if it is $\in$ \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}. Block \texttt{b} is a subblock of a \texttt{n.blocks[i]} if it is a direct subblock of it or subblock of a direct subblock of it. Block b is direct superblock of block c if c is direct subblock of b.
The set of operations of block $b$ are the operations in the leaf sub blocks of $b$. We show the set of operations of block by \nf{ops(b)}. We also say $b$ contains $op$ if $op\in ops(b)$.
\end{definition}

@@

\begin{definition}
 A block \nf{b} in \nf{n.blocks} is \emph{established} at time $t$ if \nf{n.head} is greater than the index of \nf{b} in \nf{n.blocks} at time $t$. Block \nf{b} is in \emph{EST\sub{n, t}} if \nf{b} is a subblock of \nf{b$^\prime$} in \nf{n.blocks} such that \nf{b$^\prime$} is established  at time $t$.
\end{definition}

\begin{observation}\label{head}
Once a block \nf{b} is written in \nf{n.blocks[i]} then \nf{n.blocks[i]} never changes.
\end{observation}

\begin{lemma}[headProgress] \label{lem::sizeProgress}
\nf{n.head} is non-decreasing over time and \nf{n.blocks[i].end\sub{left}} $\geq$ \nf{n.blocks[i-1].end\sub{left}},\nf{n.blocks[i].end\sub{right}} $\geq$ \nf{n.blocks[i-1].end\sub{right}}.
\end{lemma}
\begin{proof}
The first claim follows trivially from the pseudocode since \nf{n.head} is only incremented. Also  when \nf{n.blocks[i]} is created its \nf{end\sub{left}, end\sub{right}} are greater than or equal to the values in \nf{n.blocks[i-1]}. Since \nf{blocks[i-1].end\sub{dir}}$<$\nf{dir.size = blocks[i].end\sub{dir}} (Lines~\ref{lastLine, prevLine}).
\end{proof}

\begin{lemma}
Every block has most one direct superblock.
\end{lemma}
\begin{proof}
To show this we are going to refer to the way \nf{n.blocks[]} is partitioned while propagating blocks up to \nf{n.parent}. \nf{n.CreateBlock(i)} merges the blocks in \nf{n.left.blocks[n.blocks[i-1].end\sub{left}..n.blocks[i].end\sub{left}]} and \nf{n.right.blocks[n.blocks[i-1].end\sub{right}..n.blocks[i].end\sub{right}]} (Lines \ref{lastLine, pr}). Since \nf{end\sub{left},end\sub{right}} are non-decreasing, so the range of the subblocks of \nf{n.blocks[i]} which is \nf{(n.blocks[i-1].end\sub{dir}+1..n.blocks[i].end\sub{dir})} does not overlap with the range of the subblocks of \nf{n.blocks[i-1]}.
\end{proof}

\begin{corollary}[No Duplicates]\label{append}
If \nf{op} is in \nf{n.blocks[i]} then there is no \nf{j$\neq$i} such that \nf{op}$\in$\nf{ops(n.blocks[j])}.
\end{corollary}
%\begin{proof}
%\nf{Append(op)} adds \nf{op} to \nf{l\textsubscript{p}.blocks}(Line~\ref{addOP}) and \nf{Propagate()} recursively propagates \nf{op} up to the root.
%By lemma \ref{doublyRefresh} we know that operation \nf{op} propagates from child to parent at each level.
%\end{proof}

\begin{invariant}[headPosition] \label{lem::headPosition} If the value of \nf{n.head} is \nf{h} then, \nf{n.blocks[i]=null} for \nf{i>h} and \nf{n.blocks[i]$\neq$null} for \nf{i<h}.
\end{invariant}
\begin{proof}
The invariant is true initially since 1 is assigned to \nf{n.head} and \nf{n.blocks[x]} is null for every \nf{x}. The truth of the invariant may be affected by writing into \nf{n.blocks} or incrementing \nf{n.head}.

Some value is written into \nf{n.blocks[head]} only in Line 313. It is obvious that writing into \nf{n.blocks[head]} presereves the invariant. The value of \nf{n.head} is modified only in lines \ref{incrementHead1}], \ref{incrementHead2}. Depending on wether the \nf{TryAppend()}in Line\ref{cas} succeeded or not we show that the claim holds after the increment lines of \nf{n.head} in either case. If \nf{head} is incremented to $h$ it is sufficient to show \nf{n.blocks[h]}$\neq$\nf{null} to prove the invariant still holds. In the first case the process applied a successful \nf{TryAppend(new,h)} in line \ref{okcas}, which means \nf{n.blocks[h]} is not null anymore. Note that wether\ref{incrementHead1} returns true or false after Line \nf{n.head} we know has been incremented from Line \ref{readSize}. The failure case is also the same since it means some value is written into \nf{n.blocks[head]} by some process.
%At the end of every \nf{n.Refresh()} with a block  with size greater than 0 returned by \nf{CreateBlock()} \nf{n.head} is incremented (Lines \ref{incrementHead1}, \ref{incrementHead2}). If a process went to sleep before incrementing the \nf{established} (line \ref{incrementHead1}), nothing can be appended to \nf{n.blocks} until another process increments \nf{n.establsihed} (Line \ref{incrementHead2}).
\end{proof}
\it{Explain More}

@@

\begin{lemma}[establishedOrder]\label{lem::establishedOrder}
  If  time $t<$ time $t^\prime$, then \nf{ops(EST\sub{n, t})}$\subseteq$ \nf{ops(EST\sub{n, t$^\prime$})}.
\end{lemma}
\begin{proof}
Blocks are only appended (not modified) with CAS to \nf{n.blocks[n.head]} and \nf{n.head} is non-decreasing, so the set of operations in established blocks of a node can only grows.
\end{proof}

@@

\nf{CreateBlock()} aggregates the blocks in the children that are not already established in the parent into one block. If a \nf{Refresh()} procedure returns true it means it has appended the block created by \nf{CreateBlock()} into the parent node's sequence. So suppose two \nf{Refresh}es fail. Since the first \nf{Refresh()} was not successful, it means another CAS operation by a \nf{Refresh}, concurrent to the first \nf{Refresh()}, was successful before the second \nf{Refresh()}. So it means the second failed \nf{Refresh} is concurrent with another successful \nf{Refresh()} that assuredly has read block before the mentioned line \nf{35}. After all it means if any of the \nf{Refresh()} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.

\begin{lemma}[trueRefresh] \label{lem::trueRefresh}
  Let $t_i$ be the time a \nf{n.Refresh()} operation is invoked and $t_t$ be the time it terminates. Suppose \nf{n.Refresh()}'s \nf{TryAppend(new, s)} returns \nf{true}, then \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} $\subseteq$ \nf{ops(EST\sub{n, t\sub{t}})}.
\end{lemma}
\begin{proof}
From Lemma \ref{lem::establishedOrder} we know that \nf{ops(EST\sub{n, t\sub{i}})}$\subseteq$\nf{ops(EST\sub{n, t\sub{t}})}. So it remains to show the operations of \nf{ops(EST\sub{n.left, t\sub{i}})} $\cup$ \nf{ops(EST\sub{n.right, t\sub{i}})} - \nf{ops(EST\sub{n, t\sub{i}})}, which we call new operations, are all in \nf{ops(EST\sub{n, t\sub{t}})}. If \nf{TryAppend} returns true a block is appended to \nf{n}. From the code of the \nf{CreateBlock()} this block includes the established blocks in n's children at $t_i$. Since the head in \nf{CreateBlock()} is read after $t_i$. So the new operations are in the block appended to $n$.
